---
title: "Impact of Correlation on Concordance"
format:
  html:
    theme: sandstone
    embed-resources: true
    code-fold: true
    toc: true
warning: false
vignette: >
  %\VignetteIndexEntry{Impact of Correlation on Concordance}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: references.bib
---

```{r}
#| label: setup

# Load packages needed for this report:
library(demcrit)
library(tidyverse) # dor data wrangling
library(MASS)      # for mvrnorm()
library(caret)     # for confusionMatrix()
library(psych)     # for cohen.kappa()
library(gt)        # for table formatting
```

During the article's review process, the following question arose:

> How can **correlation** between FAQ scores and neuropsychological measures **inflate** concordance estimates?

Because this question is not a straightforward one to address conceptually, we conducted a brief simulation experiment to explore it empirically.

This vignette presents the R code and results of a simulation study comparing algorithms’ concordance and PDD rate estimates in a simplified setting. In this scenario, the only differences between algorithms are the operationalisations of IADL deficit (FAQ \> 7 vs. FAQ item 9 \< 1) and global cognitive deficit (MMSE \< 26 vs. MoCA \< 27), while all other diagnostic criteria are assumed to be satisfied.

Crucially, **within algorithms**, we vary correlation between FAQ and MMSE/MoCA in the data-generating process, allowing us to compare estimates under two conditions: one with no correlation (**independent** scenario) and one with a correlation of the size consistent with that observed in our data (**correlated** scenario, see @tbl-data).

```{r}
#| label: tbl-data
#| tbl-cap: Distribution moments (means and standard deviations) and Pearson's correlations between FAQ, MMSE and MoCA as measured in the observed data.
#| echo: false

data.frame(
  var = c("FAQ", "MMSE", "MoCA"),
    M = c(4.05, 26.69, 24.07),
    SD = c(4.89, 2.22, 3.48),
    cor_FAQ = c(1, -0.21, -0.27),
    cor_MMSE = c(-.21, 1, 0.63),
    cor_MoCA = c(-0.27, 0.63, 1)
) |>
  gt_apa_table() |>
  tab_spanner(label = "Moments", columns = c(M, SD), gather = FALSE) |>
  tab_spanner(label = "Correlations", columns = starts_with("cor"), gather = FALSE) |>
  cols_label(var ~ "") |>
  cols_label_with(columns = starts_with("cor"), \(x) stringr::str_replace(x, "cor_", ""))
```

## Data simulation

```{r}
#| label: simulation-parameters

# Simulation parameters:
n <- 2000 # sample size high enough to get stable estimates
k <- 2000 # enough iterations to see trends 

# Moments of the multivariate normal distribution:
mu <- c(FAQ = 4.05, MMSE = 26.69, MoCA = 24.07) # vector of means
sigma <- c(FAQ = 4.89, MMSE = 2.22, MoCA = 3.48) # vector of standard deviations

# Correlation matrix from observed data:
corrs <- matrix(
  c(1, -0.21, -0.27,
    -0.21, 1, 0.63,
    -0.27, 0.63, 1),
  nrow = 3,
  dimnames = list(
    c("FAQ", "MMSE", "MoCA"),
    c("FAQ", "MMSE", "MoCA")
  )
)

# Get variance-covariance matrix for the correlated case:
sigma_corr <- cor2cov(corrs, sigma)

# Set FAQ/cognition covariances to zero for the independent case:
sigma_indep <- sigma_corr
sigma_indep["FAQ", ] <- c(sigma_indep["FAQ", "FAQ"], 0, 0)
sigma_indep[, "FAQ"] <- c(sigma_indep["FAQ", "FAQ"], 0, 0)
```

### FAQ, MMSE & MoCA

To approximate the situation examined in the study, we simulate FAQ, MMSE and MoCA data from a common multivariate. In other words, each subject $i$ was assumed to have the following vector of latent scores:

$$
 Y_i = \begin{bmatrix} FAQ_i \\ MMSE_i \\ MoCA_i \end{bmatrix} \sim \mathcal{MVN_3}(\boldsymbol{\mu}, \boldsymbol{\Sigma})
$$ {#eq-generate}

where $\boldsymbol\mu$ denotes the vector of means and $\boldsymbol\Sigma$ the variance–covariance matrix. Across all simulations, the vector of means was fixed based on observed data (see @tbl-data):

$$
\boldsymbol\mu = \begin{bmatrix} 4.05 \\ 26.69 \\ 24.07 \end{bmatrix}
$$ {#eq-mu}

Two different specification of the variance-covariance matrix were used. The first, based on observed data (@tbl-data), represents the **correlated** scenario:

$$
\boldsymbol\Sigma_{corr} =
  \begin{bmatrix} 23.91 & -2.28 & -4.59 \\
                  -2.28 &  4.93 & 4.87  \\
                  -4.59 &  4.87 & 24.07
  \end{bmatrix}
$$ {#eq-sigma-corr}

To simulate the **independent** scenario, covariance between FAQ and MMSE/MoCA was set to zero, yielding:

$$
\boldsymbol\Sigma_{indep} =
  \begin{bmatrix} 23.91 &  0    & 0    \\
                   0    &  4.93 & 4.87  \\
                   0    &  4.87 & 24.07
  \end{bmatrix}
$$ {#eq-sigma-indep}

### FAQ item 9
To simulate data for FAQ item 9, we assumed $\boldsymbol{\tau}$**-equivalence** of FAQ (i.e., that all items share the same true score, cf. @trizano-hermosilla2016). We then applied probit link to the FAQ total score of each simulated patient to obtain the probability of a positive response to any item:

$$
p_i = \phi(z_i)
$$ {#eq-p}

where $\phi(\cdot)$ is the standard normal cumulative distribution function and $z_i$ is the standardised FAQ score for simulated patient $i$. The observed FAQ item 9 score was subsequently drawn from a binomial distribution via:

$$
\text{FAQ item 9}_i \sim \mathcal{Binomial}(n = 4, p_i)
$$ {#eq-faq9}

If the simulated FAQ item 9 score exceeded the corresponding simulated FAQ total score, it was set at zero. Finally, all simulated FAQ total, MMSE and MoCA scores were left-censored at 0 and right-censored at 30 to reflect real-world range of the respective scales.

```{r}
#| label: simulation-func

# Data-generating function:
simulate_pdd <- function(N, Mu, Sigma) {
  # Generate FAQ, MMSE and MoCA latent data:
  dat <- as.data.frame(MASS::mvrnorm(N, mu = Mu, Sigma = Sigma))
  # Generate FAQ item 9 data:
  p <- pnorm(dat$FAQ, mean = Mu["FAQ"], sd = sqrt(Sigma["FAQ", "FAQ"]))
  dat$FAQ9 <- rbinom(N, size = 4, prob = p)
  dat$FAQ9 <- ifelse(dat$FAQ9 > dat$FAQ, 0, dat$FAQ9)
  # Bound scores realistically:
  dat$FAQ  <- pmin(pmax(round(dat$FAQ), 0), 30)
  dat$MMSE <- pmin(pmax(round(dat$MMSE), 0), 30)
  dat$MoCA <- pmin(pmax(round(dat$MoCA), 0), 30)
  # Define PDD classifications:
  dat$A1 <- as.numeric(dat$FAQ > 7 & dat$MMSE < 26)
  dat$A2 <- as.numeric((dat$FAQ9 > 1) & dat$MMSE < 26)
  dat$A3 <- as.numeric(dat$FAQ > 7 & dat$MoCA < 27)
  dat$A4 <- as.numeric((dat$FAQ9 > 1) & dat$MoCA < 27)
  # Return the data:
  dat
}
```

## Diagnostic algorithms

After generating the data using the process described above, each simulated patient was classified as suffering from PDD (or not) according to all of the following four algorithms:

$$
\begin{aligned}
  A1_i &= \mathbb{1}\{\text{FAQ}_i > 7 \ \wedge\ \text{MMSE}_i < 26\}, \\
  A2_i &= \mathbb{1}\{\text{FAQ item 9}_i > 1 \ \wedge\ \text{MMSE}_i < 26\}, \\
  A3_i &= \mathbb{1}\{\text{FAQ}_i > 7 \ \wedge\ \text{MoCA}_i < 27\}, \\
  A4_i &= \mathbb{1}\{\text{FAQ item 9}_i > 1 \ \wedge\ \text{MoCA}_i < 27\}
\end{aligned}
$$ {#eq-algo}

where $\mathbb{1}\{\cdot\}$ denotes the indicator function. In other words, simulated cases fulfilling both conditions in parentheses were classified as having PDD (PDD = 1); all others were classified as non-PDD (PDD = 0).

### Data

```{r}
#| label: data

d0 <- simulate_pdd(n, mu, sigma_indep) # independent scenario
d1 <- simulate_pdd(n, mu, sigma_corr)  # correlated scenario
```

To ensure reliable results, N = `r n` cases were generated in this experiment. The simulation was based on the observed means, standard deviations and correlations as specified above. The **independent variable** in this experiment was presence versus absence of correlation between FAQ and cognitive scores. Accordingly, the simulation was conducted under two conditions:

* **Independent** scenario using $\boldsymbol\Sigma_{indep}$ (@eq-sigma-indep),
* **Correlated** scenario using $\boldsymbol\Sigma_{corr}$ (@eq-sigma-corr).

All other simulation parameters (i.e., the number of participants, the vector of variable means, the variances of all variables, and the covariance between MMSE and MoCA) were held constant across conditions.

## Statistical analysis

```{r}
#| label: compare-func

# Function for comparing two algorithms within the same scenario:
compare_algos <- function(dat, a1, a2) {
  x <- dat[[a1]]
  y <- dat[[a2]]
  cm <- caret::confusionMatrix(factor(x), factor(y), positive = "1")
  k <- psych::cohen.kappa(cbind(x, y))$kappa
  data.frame(
    Accuracy = cm$overall["Accuracy"],
    BalAcc = cm$byClass["Balanced Accuracy"],
    Kappa = k
  )
}

# Function for results summary:
summarise_results <- function(d0, d1, r = 1) {
  # PDD rates:
  rates <- map_dfr(paste0("A", 1:4), function(i) {
    c(indep = mean(d0[[i]]),
      corrs = mean(d1[[i]])
      )
  }) |>
    mutate(diff = corrs - indep, rep  = r) |>
    rownames_to_column("algo")
  # Concordance:
  conc <- expand.grid(
    ref = paste0("A", 1:4),
    pred = paste0("A", 1:4),
    scenario = c("indep", "corrs")
  )
  for (i in c("Accuracy", "BalAcc", "Kappa")) {
    conc[[i]] <- NA
  }
  for (i in seq_len(nrow(conc))) {
    y <- as.character(conc[i, "ref"])
    x <- as.character(conc[i, "pred"])
    if (conc[i, "scenario"] == "indep") {
      conc[i , c("Accuracy", "BalAcc", "Kappa")] <- compare_algos(d0, x, y)
    } else if (conc[i, "scenario"] == "corrs") {
      conc[i, c("Accuracy", "BalAcc", "Kappa")] <- compare_algos(d1, x, y)
    }
  }
  conc$rep <- r
  # Return the results:
  list(rates = rates, concordance = as_tibble(conc))
}
```

After we generated the data, the following statistical estimates were computed within each scenario (i.e., independent and correlated):

* **PDD rate** for each algorithm,
* **accuracy** for each pair of algorithms,
* **balanced accuracy** for each pair of algorithms,
* **Cohen's** $\boldsymbol\kappa$ for each pair of algorithms.

This procedure was repeated for `r k` iterations. Subsequently, the distributions of PDD rates and concordance measures were compared between the same (pairs of) algorithms across the two scenarios. Independent samples *t*-tests were used to compute 95% confidence intervals for the differences between scenarios within each diagnostic algorithm.

## Results

```{r}
#| label: analysis
#| cache: true

# Calculate all results:
res <- lapply(seq_len(k), function(i) {
  d0 <- simulate_pdd(n, mu, sigma_indep)
  d1 <- simulate_pdd(n, mu, sigma_corr)
  summarise_results(d0, d1, i)
})

# Extract PDD rates and concordance separately:
res_rates <- map_dfr(seq_len(k), \(i) res[[i]]$rates)
res_conc <- map_dfr(seq_len(k), \(i) res[[i]]$conc)

# Conduct t-tests for PDD rates:
stats_rates <- res_rates |>
  pivot_longer(c(indep, corrs), names_to = "scenario", values_to = "rate") |>
  group_by(algo) |>
  rstatix::t_test(rate ~ scenario, detailed = TRUE, var.equal = TRUE) |>
  mutate(
    Algorithm = glue::glue("A{algo}"),
    Independent = estimate2,
    Correlated = estimate1,
    Difference = estimate,
    Low = conf.low,
    High = conf.high
  ) |>
  dplyr::select(Algorithm, Independent, Correlated, Difference, Low, High)

# Conduct t-tests for concordance metrics:
stats_conc <- map_dfr(c("Accuracy", "BalAcc", "Kappa"), function(y) {
  res_conc |>
    filter(ref != pred) |>
    mutate(Comparison = glue::glue("{ref}{pred}")) |>
    group_by(Comparison) |>
    rstatix::t_test(formula(glue::glue("{y} ~ scenario")), detailed = TRUE, var.equal = TRUE) |>
    mutate(
      Index = y,
      Independent = estimate1,
      Correlated = estimate2,
      Difference = estimate,
      Low = conf.low,
      High = conf.high,
      Type = case_when(
        Comparison %in% c("A1A3", "A3A1", "A2A4", "A4A2") ~ "Aligned IADL & Mismatched Global Cognition",
        Comparison %in% c("A1A2", "A2A1", "A3A4", "A4A3") ~ "Mismatched IADL & Aligned Global Cognition",
        Comparison %in% c("A1A4", "A4A1", "A2A3", "A3A2") ~ "Mismatched IADL & Mismatched Global Cognition"
      )
    ) |>
    dplyr::select(Index, Type, Comparison, Independent, Correlated, Difference, Low, High)
})
```

The expected values (denoted $E(\cdot)$) and the differences between the independent and correlated scenarios in PDD rates, accuracy, balanced accuracy and Cohen's $\kappa$ are presented in @tbl-rates, @tbl-acc, @tbl-balacc, and @tbl-kappa, respectively.

Across all algorithms, the **correlated** scenario generated **higher PDD rates** than the independent scenario, with differences of approximately 2–3% (@tbl-rates).

```{r}
#| label: tbl-rates
#| tbl-cap: Expectations of PDD rates and differences between the independent and correlated scenarios based on independent sample t-tests.
#| echo: false

stats_rates|>
  gt_apa_table() |>
  tab_spanner(label = "95% CI", columns = c(Low, High), gather = FALSE) |>
  tab_spanner(label = "Rate Difference", columns = c(Difference, Low, High), gather = FALSE) |>
  tab_spanner(label = md("*E*(PDD Rate)"), columns = c(Independent, Correlated), gather = FALSE) |>
  fmt_number(decimals = 3)
```

```{r}
#| label: tab-func
#| echo: false

make_concordance_table <- function(dat, index, label) {
  dat |>
    filter(Index == index) |>
    dplyr::select(-Index) |>
    #gt_apa_table(grp = "Type") |>
    gt(group = "Type") |>
    tab_spanner(label = "95% CI", columns = c(Low, High), gather = FALSE) |>
    tab_spanner(label = md(glue::glue("{label} difference")), columns = c(Difference, Low, High), gather = FALSE) |>
    tab_spanner(label = md(glue::glue("*E*({label})")), columns = c(Independent, Correlated), gather = FALSE) |>
  fmt_number(decimals = 3)
}
```

The effect of the data-generating scenario on **concordance indeces** was more nuanced. For **accuracy**, the **correlated** scenario generally yielded **slightly lower** estimates than the independent one across most algorithm pairs (@tbl-acc). However, the magnitude of these differences did not exceed 1%.

```{r}
#| label: tbl-acc
#| tbl-cap: Expectations of accuracy estimates and differences between the independent and correlated scenarios based on independent sample t-tests.
#| echo: false

make_concordance_table(stats_conc, "Accuracy", "Accuracy")
```

Results for **balanced accuracy** were more heterogeneous (@tbl-balacc). In algorithm pairs where **both** the IADL and global cognition criteria were **mismatched** (e.g., A1 vs A4), balanced accuracy was consistently **higher** under the **correlated** scenario, by roughly 3%.

When the only **mismatch** concerned the **IADL** criterion, the **direction** of difference in balanced accuracy reversed depending on which algorithm was used as the reference (e.g., when A1 predicted A2, the correlated scenario yielded lower balanced accuracy, whereas when A2 predicted A1, it yielded higher values). The magnitude of this effect was comparable in both directions.

Finally, when the only **mismatch** involved **global cognition** criterion, the **correlated** scenario led to **higher balanced accuracy** estimates in three out of four cases. The maximum difference was similar to the other comparison types, reaching up to 3.8%.

```{r}
#| label: tbl-balacc
#| tbl-cap: Expectations of balanced accuracy estimates and differences between the independent and correlated scenarios based on independent sample t-tests.
#| echo: false

make_concordance_table(stats_conc, "BalAcc", "Balanced accuracy")
```

Lastly, Cohen's $\kappa$ showed the clearest and most consistent pattern of results (@tbl-kappa). In all cases, the **correlated** scenario yielded **higher** $\kappa$ values than the independent one, with differences ranging from `r do_summary(min(abs(subset(stats_conc, Index == "Kappa")$Difference)), 3)` to `r do_summary(max(abs(subset(stats_conc, Index == "Kappa")$Difference)), 3)`.

```{r}
#| label: tbl-kappa
#| tbl-cap: Expectations of Cohen's κ estimates and differences between the independent and correlated scenarios based on independent sample t-tests.
#| echo: false

make_concordance_table(stats_conc, "Kappa", "Cohen's $\\kappa$")
```

## Conclusions

::: {.callout-tip appearance="simple" icon=false}

Based on the simulations presented in this vignette, we can conclude that the correlations between FAQ and global cognition measures of the magnitude observed in our empirical data imply that:

- the **correlated** scenario is more **sensitive** to PDD, consistently yielding **higher PDD rates**,
- the **correlated** scenario produces **accuracy estimates** that are **similar** to those from the independent scenario,
- when considering **balanced accuracy**, the **correlated** scenario diverges from the independent one in more **nuanced** ways, yielding **higher** estimates in some comparisons and **lower** in others,
- relative to the independent scenario, the **correlated** scenario tends to **slightly overestimate** Cohen’s $\kappa$ to a small degree^[Following the reference in the main text [@mchugh2012], a qualitative shift in the interpretation of Cohen’s $\kappa$ would require a difference of approximately 0.2, more than twice the largest difference observed in this simulation study.], and
- overall, the differences between the **correlated** and **independent** scenarios are likely to be **small in magnitude**, unless higher correlations between diagnostic criteria are introduced in the data-generating process.^[Although this simulation employed a correlation matrix based on our empirical data, the code can be easily modified to generate data with stronger correlations, which would, in turn, produce larger differences between scenarios. For brevity, we leave this exploration to the reader.]

:::
