---
title: "Use #1: Targets Pipeline"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Use #1: Targets Pipeline}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

> The full use of the *demcrit* package relies on patient data that cannot be shared publicly for privacy reasons. However, the package can be executed on the data owner’s computer ([Laboratory of Neurosychology](https://www.vfn.cz/en/pacienti/kliniky-ustavy/neurologicka-klinika/odborne-ambulance/#) at the First Faculty of Medicine, Charles University, and the General University Hospital in Prague).

## Set-up

Before starting, ensure that you have a local installation of the [demcrit](https://github.com/josefmana/demcrit.git) R package. If not, see the [package overview](../index.html) for installation instructions, or visit the [package website](https://josefmana.github.io/demcrit/) for function documentation.

Load the package if you wish to access the documentation directly in RStudio:

```{r}
#| label: setup
#| eval: false

library(demcrit)
```

Next, clone the [*demcrit* repository](https://github.com/josefmana/demcrit.git) to a local folder of your choice (here referred to as `studyfolder`), and install the necessary dependencies:

```{r}
#| label: deps
#| eval: false

devtools::install_deps()
```

Ensure that all required data files are present in the `studyfolder/raw-data` directory.

## Data Checking

The pipeline first runs the `data_paths()` function, which **checks for missing files** and **naming inconsistencies**:

```{r}
#| label: data-paths
#| eval: false

help("data_paths")
```

Additional functions help prevent data-entry errors or structural inconsistencies before analysis:

```{r}
#| label: other-helpers
#| eval: false

help("check_compatibility") # Looks for incompatibilities between raw item data specific for this project and REDCap data from a larger database.
help("check_names") # Looks for inconsistencies in patients names in the raw item data compare to REDCap databes.
help("check_ranges") # Looks for impossible test scores.
```

## Running the Pipeline

Finally, make sure that the [apaquarto](https://github.com/wjschne/apaquarto.git) extension (version 4.4.1) is installed in the `studyfolder/_manuscript` directory.

You can then run the full pipeline (including manuscript writing) via:

```{r}
#| label: tagets-pipeline
#| eval: false

#install.packages("targets")
#install.packages("tarchetypes")
targets::tar_make()
```

To visualize the workflow, use:

```{r}
#| label: tagets-flow
#| eval: false

targets::tar_visnetwork() # full workflow
targets::tar_glimpse() # simplified workflow (targets only)
```
