---
title: "Use #2: Out of Pipeline"
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Use #2: Out of Pipeline}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

> If you do not have access to the raw data used in our article, you can still apply the key *demcrit* functions to your own, similarly structured dataset. In this vignette, we show how to do that using a small reproducible example.

Before you start, make sure you have a local installation of the [demcrit](https://github.com/josefmana/demcrit.git) R package. If not, see the [package overview](../index.html) for installation instructions.

## Setup

Start by loading the required packages:

```{r}
#| label: setup

library(demcrit)
library(gt) # for nicely formatted tables
library(tidyverse) |> suppressPackageStartupMessages() # For data wrangling
```

## Data Simulation

Because the original dataset cannot be shared, we will simulate a similar dataset using built-in functions:

```{r}
#| label: simulate-data

set.seed(87542) # for reproducibility
data <- simulate_pdd_data()
```

You can inspect the default parameters used in this simulation via:

```{r}
#| label: show-defaults

prepare_defaults()
```

This produces the following example dataset:

```{r}
#| label: show-data
#| echo: false

data
```

Key variables include:

- `id` -- patient identifier
- `type` -- diagnostic algorithm used
- `PDD` -- diagnosis result for each patient–algorithm pair

For more details about how the data are generated, see:

```{r}
#| label: data-helps
#| eval: false

help("simulate_pdd_data") # details on the data generation algorithm
help("prepare_defaults") # details on parameter defaults
```

## Dementia Rates

To compute dementia (PDD) rates, we first prepare a specification of diagnostic algorithms and variable mappings. Then, we pass these to the `summarise_rates()` function:

```{r}
#| labels: rates

# Extract default parameters:
pars <- prepare_defaults()

# Define algorithm specifications:
algos <- with(pars, {data.frame(
  type = rownames(crits),
  glob = crits$cognition, glob_t = crits$cognition_thres,
  iadl = crits$IADL, iadl_t = crits$IADL_thres,
  atte = "", atte_t = NA,
  exec = "", exec_t = NA,
  memo = "", memo_t = NA,
  cons = "", cons_t = NA,
  lang = "", lang_t = NA
)})

# Define variable names:
vars <- data.frame(
  name = c(unique(algos$glob), unique(algos$iadl)),
  label = c(unique(algos$glob), unique(algos$iadl)),
  type = "continuous"
)

# Compute PDD rates:
rates <- summarise_rates(
  list(PDD = data, algorithms = algos),
  vars = vars,
  plot = FALSE # Setting plot to FALSE bacause it is calibrated the data from the study..
)

# Format results:
tab1 <- rates$table |>
  select(type, Global, IADL, N, Rate) |>
  gt_apa_table() |>
  cols_label(
    type ~ "Algorithm",
    Global ~ "Cognitive deficit",
    IADL ~ "IADL deficit"
  )
```

```{r}
#| label: tbl-rates
#| echo: false
#| tbl-cap: An example table showing simulated PDD rates.

tab1
```

## Pairwise Concordance

Next, we can compute pairwise concordance statistics across algorithms:

```{r}
#| label: concordances
#| warning: false

concs <- describe_concordance(list(PDD = data, algorithms = algos))
```

For details about the resulting object, see:

```{r}
#| label: conc-docs
#| warning: false

help("describe_concordance")
help("concords") # describes the concordance statistics table
```

We can visualize concordance matrices for several metrics:

```{r}
#| label: fig-concordances
#| echo: false
#| warning: false
#| layout-ncol: 2
#| fig-cap: Concordance statistic matrices
#| fig-subcap:
#|    - Cohen's kappa
#|    - Raw accuracy
#|    - Sensitivity
#|    - Specificity

concs$plots$Kappa
concs$plots$Accuracy
concs$plots$Sensitivity
concs$plots$Specificity
```

Alternatively, we can extract specific concordance statistics and present them in a table.
Table @tbl-concordance shows one such example, with algorithms sorted by raw accuracy in predicting *MoCA (1)*.

```{r}
#| label: concordance-table

tab2 <- concs$table |>
  filter(reference == "MoCA (1)") |>
  filter(reference != predictor) |>
  arrange(desc(Accuracy_raw)) |>
  select(predictor, Kappa, Accuracy, NoInformationRate_raw, AccuracyPValue) |>
  mutate(AccuracyPValue = do_summary(AccuracyPValue, 3, "p")) |>
  gt_apa_table() |>
  fmt_number(decimals = 2) |>
  cols_label(
    predictor ~ "Predictor",
    NoInformationRate_raw ~ "NIR",
    AccuracyPValue ~ "p-value"
  )
```

```{r}
#| label: tbl-concordance
#| echo: false
#| tbl-cap: Example table showing pairwise concordance statistics for five algorithms sorted by raw accuracy in predicting MoCA (1). NIR stands for *No Information Rate*, and p-values correspond to one-sided binomial tests of Accuracy > NIR.

tab2
```
